Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-13 numerical-mhd-ct.tex
I still managed to mess up the cosmological B field.  Ok this time I have it right, really!  In particular, the paper now adopts the MHD-CT B definitions throughout and explicits lists the changes in the MHD-Dedner scheme.

Cosmological MHD can be treated in a number of ways.  As described in 
the previous section, and equation~\ref{eq:induction} the MUSCL with Dedner
method treats the cosmological dilution of
$\vecB$ with a source term,  However, one can eliminate the right hand side of equation 
\ref{eq:induction} by defining
$\vecB^\prime = \vecB / \sqrt{a}.$
Then equation~\ref{eq:induction} becomes
$$ 
\frac{\partial \vecB^\prime}{\partial t} - \frac{1}{a}  \curl (\vecv \times
\vecB^\prime)  = 0, \label{eq:induction_cosmo_prime_nosource}
$$
and we can reuse each part of the non-cosmological algorithm (hyperbolic solver,
electric field creation, interpolation, projection, and flux correction) without
modification.

>>>>

%Cosmological MHD can be treated in a number of ways.  As described in 
%the previous section, and equation~\ref{eq:induction} the MUSCL with Dedner
%method treats the cosmological dilution of
%$\vecB$ with a source term,  However, one can eliminate the right hand side of equation 
%\ref{eq:induction} by defining
%$\vecB^\prime = \vecB / \sqrt{a}.$
%Then equation~\ref{eq:induction} becomes
%$$ 
%\frac{\partial \vecB^\prime}{\partial t} - \frac{1}{a}  \curl (\vecv \times
%\vecB^\prime)  = 0, \label{eq:induction_cosmo_prime_nosource}
%$$
%and we can reuse each part of the non-cosmological algorithm (hyperbolic solver,
%electric field creation, interpolation, projection, and flux correction) without
%modification.  
%GB: I removed the above because the rest of the paper now uses the definition
% of B in MHD-CT.

### 2013-06-13 numerical-mhd-dedner.tex
I still managed to mess up the cosmological B field.  Ok this time I have it right, really!  In particular, the paper now adopts the MHD-CT B definitions throughout and explicits lists the changes in the MHD-Dedner scheme.

\subsection{Hydrodynamics: MUSCL}
\label{sec.num.hydro-muscl}

@@

>>>>

\subsection{Hydrodynamics and Magnetohydrodynamics: MUSCL with Dedner cleaning}
\label{sec.num.hydro-muscl}

@@

Finally, we note that for cosmological simulations, this solver uses a slightly different definition of the magnetic field than used in the rest of the paper.  In particular, the field is defined as $\vecB = a^{3/2} \vecB^{\prime}$ (where $\vecB^{\prime}$ is the proper field strength).  This adds a source term of $-\dot{a}/(2a) \vecB$ on the right-hand side of equation~\ref{eq:induction}, removes the factor of $a$ in the $\vecB$ term in the energy equation (eq.~\ref{eq:total_energy_def}) and in the definition of the isotropic pressure $p^*$.

### 2013-06-13 overview.tex
I still managed to mess up the cosmological B field.  Ok this time I have it right, really!  In particular, the paper now adopts the MHD-CT B definitions throughout and explicits lists the changes in the MHD-Dedner scheme.

\begin{eqnarray} 
  \frac{\partial \rho}{\partial t} 
  + \frac{1}{a} \div (\rho \vecv) & = & 0,
  \label{eq:mass} \\
%
  \frac{\partial \rho \vecv}{\partial t}  
  + \frac{1}{a} \div \left(\rho \vecv \vecv + \myvec{I}p^* - 
    \frac{\vecB \vecB}{a} \right) & = &
  - \frac{\dot{a}}{a} \rho \vecv - \frac{1}{a} \rho \grad \phi,
% 
%          + \frac{1}{a} ( \vecv \cdot \grad ) \vecv
%    & = & - \frac{\dot{a}}{a} \vecv 
%          - \frac{1}{a \rho} \grad p
%          - \frac{1}{a \rho} (\nalba \times \vecB) \times \vecB
%          - \frac{1}{a} \grad \phi, 
  \label{eq:momentum} \\
%
  \frac{\partial E}{\partial t} 
  + \frac{1}{a} \div \left[ (E + p^*) \vecv - 
    \frac{1}{a} \vecB (\vecB \cdot \vecv) \right] & = &
  - \frac{\dot{a}}{a} \left( 2E - \frac{B^2}{2a} \right) - 
  \frac{1}{a} \vecv \cdot \grad \phi 
  - \Lambda + \Gamma + \frac{1}{a^2} \div \fcond,
% 
%         + \frac{1}{a} \vecv \cdot \grad E
%  & = & - \frac{\dot{a}}{a} (3 \frac{p}{\rho} + {\vecv \cdot \vecv} )
%       - \frac{1}{a \rho } \div (p \vecv) 
%%\nonumber   & &   %uncomment for 2-column
%        - \frac{1}{a} \vecv \cdot \grad \phi 
%               + \Gamma - \Lambda.
  \label{eq:total_energy}  \\
%
  \frac{\partial \vecB}{\partial t} - 
  \frac{1}{a}  \grad \times (\vecv \times \vecB) & = & 
  -\frac{\dot{a}}{2 a} \vecB. \label{eq:induction}
\end{eqnarray}

@@

\begin{eqnarray}
\vecx & = & \vecx' / a, \\
\vecv & = & a \hspace{0.5mm} \dot{\vecx} = 
              \vecvp - \dot{a}\vecx, \\
\rho  & = & a^3 \rhop,   \\
p     & = & a^3 \pp, \\
E     & = & a^3 \left(\Ep - 
            \dot{a} \, \vecx \cdot \vecvp - 
            \frac{1}{2} \dot{a}^2 \vecx^2\right), \\
\phi  & = & \phip + \frac{1}{2} \, a \ddot{a} \, \vecx^2, \\
\vecB & = & a^2 \vecB^\prime,
\end{eqnarray}
where primes indicate quantities in proper coordinates (fixed frame).
We note that the definition of comoving $\vecB$ used here is such
that a uniform field is constant in a homogeneous, uniform universe
and is also the quantity used in the MHD-Dedner solver described
below; however, it is not universal,
and in particular is slightly different from that used both in \citet{Li08a}
and that used in the MHD-CT solver.
%These definitions result in the following comoving equivalents to eqs. 
%(\ref{eq:mass})-(\ref{eq:potential}).
%%
%
%\begin{eqnarray} 
%\frac{\partial \rhop}{\partial t} 
%          + \frac{1}{a} \vecvp \cdot \grad \rhop 
%    & = & - \frac{1}{a} \rhop \div \vecvp, 
%        \label{eq:mass_prime} \\
%%
%\frac{\partial \vecvp}{\partial t}  
%          + \frac{1}{a} ( \vecvp \cdot \grad ) \vecvp 
%    & = & - \frac{\dot{a}}{a} \vecvp 
%          - \frac{1}{a \rhop} \grad \pp 
%          - \frac{1}{a} \grad \phip, 
%        \label{eq:momentum_prime} \\
%%
%\frac{\partial \Ep}{\partial t} 
%         + \frac{1}{a} \vecvp \cdot \grad \Ep 
%   & = & - \frac{\dot{a}}{a} (3 \frac{\pp}{\rhop} + {\vecvp \cdot \vecvp} )
%         - \frac{1}{a \rhop } \div (p \vecvp) 
%%\nonumber   & &   %uncomment for 2-column
%         - \frac{1}{a} \vecvp \cdot \grad \phip 
%                + \Gamma - \Lambda.
%       \label{eq:total_energy_prime} \\
%%
%\Ep  & = & \ep + \frac{1}{2} {\vecvp \cdot \vecvp},
%        \label{eq:total_energy_def_prime} \\
%%
%\ep                  & = & 
%        \pp/\left[ \left( \gamma - 1 \right) \rhop \right],
%      \label{eq:state_prime} \\ 
%{\nabla}^2 \phip    & = & 
%        \frac{4\pi G}{a} ( \rhop_{total} - \rhop_0 ).
%      \label{eq:poisson}
%\end{eqnarray}
%%
%%We use the proper peculiar baryonic velocity $\vecv_b \equiv a(t) d\vecx / dt$, proper pressure $p$, and modified gravitational potential $ \phi $ which is related to the potential in proper coordinates $\Phi$ by 
%%$$\phi \equiv \Phi + \frac{1}{2} a \ddot{a} \vecx^2. $$
%%The density, however, is comoving:
%%$$\rho_b \equiv \rho_{b,proper}a(t)^3, $$
The expansion parameter $a \equiv 1/(1 + z)$ follows the expansion of
a smooth, homogeneous background, where $z$, the redshift, is a
function only of $t$.  All spatial derivatives are determined with
respect to the comoving position $\vecx$, which removes the universal
expansion from the coordinate system. The evolution of $a(t)$ is
governed by the second Friedmann equation for the expansion of a
spatially homogeneous and isotropic universe
%
\begin{eqnarray}
\frac{\ddot{a}}{a} & = & 
      - \frac{4 \pi G }{3 a^3 } (\rho_0 
      + 3p_0/c^2) 
      + \Lambda_c c^2 / 3 .
      \label{eq:expansion} 
\end{eqnarray}
%
Here $\rho_0$ is the mean comoving mass density (including both
baryonic and dark matter), $p_0$ is the comoving background pressure
contribution, and $\Lambda_c$ is the cosmological constant.
%$\gamma$ is the ratio of specific heats, and $\Lambda$ is the cosmological constant.
%We note that if $\gamma$ is not $5/3$ then $E$ in the first term on
%the right side of eq.~(ref{eq:total_energy}) must be replaced with 
%eq.~(\ref{eq:state}).
This system of equations is limited to the non-relativistic regime and
assumes that curvature effects are not important --- both assumptions
are reasonable as long as the size of the simulated region is small
compared to the radius of curvature and the Hubble length $c/H$, where
$c$ is the speed of light and $H = \dot{a}/a$ is the Hubble constant.

>>>>

\begin{eqnarray} 
  \frac{\partial \rho}{\partial t} 
  + \frac{1}{a} \div (\rho \vecv) & = & 0,
  \label{eq:mass} \\
%
  \frac{\partial \rho \vecv}{\partial t}  
  + \frac{1}{a} \div \left(\rho \vecv \vecv + \myvec{I}p^* - 
    \frac{\vecB \vecB}{a} \right) & = &
  - \frac{\dot{a}}{a} \rho \vecv - \frac{1}{a} \rho \grad \phi,
% 
%          + \frac{1}{a} ( \vecv \cdot \grad ) \vecv
%    & = & - \frac{\dot{a}}{a} \vecv 
%          - \frac{1}{a \rho} \grad p
%          - \frac{1}{a \rho} (\nalba \times \vecB) \times \vecB
%          - \frac{1}{a} \grad \phi, 
  \label{eq:momentum} \\
%
  \frac{\partial E}{\partial t} 
  + \frac{1}{a} \div \left[ (E + p^*) \vecv - 
    \frac{1}{a} \vecB (\vecB \cdot \vecv) \right] & = &
  - \frac{\dot{a}}{a} \left( 2E - \frac{B^2}{2a} \right) - 
  \frac{\rho}{a} \vecv \cdot \grad \phi 
  - \Lambda + \Gamma + \frac{1}{a^2} \div \fcond,
% 
%         + \frac{1}{a} \vecv \cdot \grad E
%  & = & - \frac{\dot{a}}{a} (3 \frac{p}{\rho} + {\vecv \cdot \vecv} )
%       - \frac{1}{a \rho } \div (p \vecv) 
%%\nonumber   & &   %uncomment for 2-column
%        - \frac{1}{a} \vecv \cdot \grad \phi 
%               + \Gamma - \Lambda.
  \label{eq:total_energy}  \\
%
  \frac{\partial \vecB}{\partial t} - 
  \frac{1}{a}  \grad \times (\vecv \times \vecB) & = & 
  %-\frac{\dot{a}}{2 a} \vecB. 
  0
  \label{eq:induction}
\end{eqnarray}

@@

\begin{eqnarray}
\vecx & = & \vecx' / a, \\
\vecv & = & a \hspace{0.5mm} \dot{\vecx} = 
              \vecvp - \dot{a}\vecx, \\
\rho  & = & a^3 \rhop,   \\
p     & = & a^3 \pp, \\
E     & = & a^3 \left(\Ep - 
            \dot{a} \, \vecx \cdot \vecvp - 
            \frac{1}{2} \dot{a}^2 \vecx^2\right), \\
\phi  & = & \phip + \frac{1}{2} \, a \ddot{a} \, \vecx^2, \\
\vecB & = & a^2 \vecB^\prime,
\end{eqnarray}
where primes indicate quantities in proper coordinates (fixed frame).
We note that the definition of comoving $\vecB$ used here is such
that a uniform field is constant in a homogeneous expanding universe
and is also the quantity used in the MHD-CT solver described
below; however, it is not universal,
and in particular is slightly different from that used both in \citet{Li08a}
and used in the MHD-Dedner solver (see section~\ref{sec.num.hydro-muscl}
for more details).
%These definitions result in the following comoving equivalents to eqs. 
%(\ref{eq:mass})-(\ref{eq:potential}).
%%
%
%\begin{eqnarray} 
%\frac{\partial \rhop}{\partial t} 
%          + \frac{1}{a} \vecvp \cdot \grad \rhop 
%    & = & - \frac{1}{a} \rhop \div \vecvp, 
%        \label{eq:mass_prime} \\
%%
%\frac{\partial \vecvp}{\partial t}  
%          + \frac{1}{a} ( \vecvp \cdot \grad ) \vecvp 
%    & = & - \frac{\dot{a}}{a} \vecvp 
%          - \frac{1}{a \rhop} \grad \pp 
%          - \frac{1}{a} \grad \phip, 
%        \label{eq:momentum_prime} \\
%%
%\frac{\partial \Ep}{\partial t} 
%         + \frac{1}{a} \vecvp \cdot \grad \Ep 
%   & = & - \frac{\dot{a}}{a} (3 \frac{\pp}{\rhop} + {\vecvp \cdot \vecvp} )
%         - \frac{1}{a \rhop } \div (p \vecvp) 
%%\nonumber   & &   %uncomment for 2-column
%         - \frac{1}{a} \vecvp \cdot \grad \phip 
%                + \Gamma - \Lambda.
%       \label{eq:total_energy_prime} \\
%%
%\Ep  & = & \ep + \frac{1}{2} {\vecvp \cdot \vecvp},
%        \label{eq:total_energy_def_prime} \\
%%
%\ep                  & = & 
%        \pp/\left[ \left( \gamma - 1 \right) \rhop \right],
%      \label{eq:state_prime} \\ 
%{\nabla}^2 \phip    & = & 
%        \frac{4\pi G}{a} ( \rhop_{total} - \rhop_0 ).
%      \label{eq:poisson}
%\end{eqnarray}
%%
%%We use the proper peculiar baryonic velocity $\vecv_b \equiv a(t) d\vecx / dt$, proper pressure $p$, and modified gravitational potential $ \phi $ which is related to the potential in proper coordinates $\Phi$ by 
%%$$\phi \equiv \Phi + \frac{1}{2} a \ddot{a} \vecx^2. $$
%%The density, however, is comoving:
%%$$\rho_b \equiv \rho_{b,proper}a(t)^3, $$
The expansion parameter $a \equiv 1/(1 + z)$ follows the expansion of
a smooth, homogeneous background, where $z$, the redshift, is a
function only of $t$.  All spatial derivatives are determined with
respect to the comoving position $\vecx$, which removes the universal
expansion from the coordinate system. The evolution of $a(t)$ is
governed by the second Friedmann equation for the expansion of a
spatially homogeneous and isotropic universe
%
\begin{eqnarray}
\frac{\ddot{a}}{a} & = & 
      - \frac{4 \pi G }{3 a^3 } (\rho_0 
      + 3p_0/c^2) 
      + \Lambda_c c^2 / 3 .
      \label{eq:expansion} 
\end{eqnarray}
%
Here $\rho_0$ is the mean comoving mass density (including both
baryonic and dark matter), $p_0$ is the comoving background pressure
contribution, and $\Lambda_c$ is the cosmological constant.
%$\gamma$ is the ratio of specific heats, and $\Lambda$ is the cosmological constant.
%We note that if $\gamma$ is not $5/3$ then $E$ in the first term on
%the right side of eq.~(ref{eq:total_energy}) must be replaced with 
%eq.~(\ref{eq:state}).
This system of equations is limited to the non-relativistic regime and
assumes that curvature effects are not important --- both assumptions
are reasonable as long as the size of the simulated region is small
compared to the radius of curvature and the Hubble length $c/H$, where
$c$ is the speed of light and $H = \dot{a}/a$ is the Hubble constant.
